Do we need detailed demographic data to forecast plant population responses to climate change?

### 2016-08-30 manuscript/manuscript.tex
adds PBA line edits for R1; ready for USGS

To date, most empirical population modelers have relied on
individual-level data, with few attempts to capitalize on
population-level measures. An important exception was an effort by
Taylor and Hastings (2004) to model the population growth rate of an
invasive species to identify the best strategies for invasion control.
They used a ``density-structured'' model where the state variable is a
discrete density state rather than a continuous density measure. Such
models do not require individual-level demographic data and can
adequately describe population dynamics. Building on Taylor and Hastings
(2004), Freckleton et al. (2011) showed that density-structured models
compare well to continuous models in theory, and Queenborough et al.
(2011) provide empirical evidence that density-structured models are
capable of reproducing population dynamics at landscape spatial scales,
even if some precision is lost when compared to fully continuous models.
However, previous tests of density-structured population models have yet
to assess their ability to forecast out-of-sample observations, and they
have not included environmental covariates, which are necessary to
forecast population responses to climate change.

Addressing climate change questions with models fit to population-level
data is potentially problematic. Population growth (or decline) is the
outcome of demographic processes such as survival, growth, and
recruitment that occur at the level of individual plants. Climate can
affect each demographic process in unique, potentially opposing, ways
(Dalgleish et al. 2011). These unique climate responses may be difficult
to resolve in statistical models based on population-level data where
demographic processes are not identifiable. Futhermore, models based on
aggregated data may reflect short-term effects of one vital rate more
than others whose importance may only emerge over the long-term. For
example, a one-year change in a plant species' cover or biomass might
reflect individual growth or shrinkage, whereas the long-term trajectory
of the population might be more influenced by recruitment. The same is
true for density dependence: intraspecific density depedence may act
most strongly on vital rates, like recruitment, that are difficult to
identify from population-level data. If density dependence and/or
important climate effects are missed because of the aggregation inherent
in population-level data, then population models built with such data
will make uninformative or unreliable forecasts.

@@

We fit all models using a hierarchical Bayesian approach. We focus on
the main process and the model likelihood in what follows (full model
descriptions are in the Supporting Information). For the likelihood
models, \(\textbf{y}^X\) is always the relevant vector of observations
for vital rate \emph{X} (\emph{X} = \emph{S}, \emph{G}, \emph{R}, or
\emph{P} for survival, growth, recruitment, or population growth). For
example, \(\textbf{y}^S\) is a vector of 0s and 1s indicating whether a
genet survives from \emph{t} to \emph{t+1}, or not, for all observation
years and quadrats. All model parameters are species-specific, but we
omit subscripts for species in model descriptions below to reduce visual
clutter. For brevity, we only describe models with climate covariates
included, but models without climate covariates are simply the models
described below with the climate effects removed.

@@

Our Bayesian approach to fitting the vital rate models required choosing
appropriate priors for unknown parameters and deciding which, if any, of
those priors should be hierarchical. For each species, we fit yearly
size effects and yearly intercepts hierarchically, where year-specific
coefficients were modeled with global distributions representing the
mean size effect and intercept. Quadrat random effects were also fit
hierarchically, with quadrat offsets modeled with distributions with
mean zero and a shared variance term (independent Gaussian priors).
Climate effects were modeled as independent covariates whose prior
distributions were optimized for prediction using statistical
regularization (see \textbf{Statistical regularization: Bayesian ridge
regression} below).

@@

An important feature of any forecasting model is the rate at which
forecast skill declines as the time between an observation and a
forecast increases.
\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon}
(Petchey et al. 2015). To assess the forecast horizons of our models, we
initiate the forecast model with the population state at some time
\emph{t} and make sequential forecasts of the population at times
\(t+1, t+2, \dots, t+T\) where \emph{T} is the maximum number of years
between the initial year and the final year of our observations. For
example, if we initialize the forecast model with percent cover in 1940,
we are able to make five forecasts up to the year 1945. Forecast models
are not re-initialized with observations between years. Thus, in our
current example, the model forecast for percent cover in 1941 has a
forecast horizon of one year, the forecast in 1942 has a forecast
horizon of two years, and so on. We performed these simulations for all
model types (IPM with/without climate; QBM with/without climate) using
mean parameter values for all possible initial years. For a given
forecast distance, we averaged the correlation between forecasts and
observations. Note that our forecasts for the horizon analysis are all
made using in-sample data because we used model fits from the full data
set. Nonetheless, our simulations offer insight into the differences
among model forecast horizons.
\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes}
(Petchey et al. 2015),
\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}

@@

With climate covariates included
\textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased},
but they did so at similar rates (Fig. 4). The only exception is for
\emph{Poa secunda}, where QBM forecast accuracy remained steady as the
\textcolor{blue}{temporal distance of the forecast increased}, whereas
IPM forecast accuracy declined (Fig. 4).
\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H. comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast profieciency threshold by a large margin (Fig. 4, Table S24).}

@@

Our comparison of a traditional, demographic population model without
environmental forcing (the IPM) to an equivalent model inspired by
density-structured models (the QBM) showed that, generally, IPM
forecasts of out-of-sample plant population states were no more accurate
than forecasts from the QBM (Fig. 2; `no-climate' bars). This result
differed from our expectation that the IPM would out-perform the QBM,
because of its mechanistic representation of the perennial life cycle.
Our result also confirms theoretical (Freckleton et al. 2011) and
empirical work (Taylor and Hastings 2004, Queenborough et al. 2011)
showing that density-structured models can be useful surrogates for
demographic models when the goal is to estimate or forecast population
states over large spatial extents.

We also expected the inclusion of environmental forcing to reveal more
differences between the models. Interannual variation in weather can
affect vital rates in different ways (Dalgleish et al. 2011). Thus,
estimates of climate effects on plant population growth may be biased or
non-identifiable when the underlying statistical model is fit using
population-level data that integrates over the potentially unique
climate responses of individual vital rates. We found some evidence that
the QBM failed to detect climate effects for three species (\emph{B.
gracilis}, \emph{H. comata}, and \emph{Poa secunda}), where important
climate effects were identified in the individual vital rate models but
not in the percent cover model (Fig. 3). However, for the two species
where including climate covariates increased forecast accuracy (\emph{B.
gracilis} and \emph{Poa secunda}), forecast skill (Fig. 2) and error
(Fig. S2) were equivalent between the IPM and QBM.

@@

We also expected IPM forecast accuracy to decline at a lower rate than
the QBM as the
\textcolor{blue}{time between the model initialization and the forecast increased.}
In principle, more mechanistic models should produce better predictions,
especially under novel conditions (Evans 2012, Schindler and Hilborn
2015). In our case, the IPM explicitly models the influence of weather
on recruitment and survival, effects that may be poorly represented in
the QBM because recruitment and survival mainly affect small plants that
contribute little to year-to-year changes in percent cover. Over time,
the addition and subtraction of small plants can have large effects on
population growth, thus, explicitly modeling these effects could
contribute to a longer forecast horizon. However, we found no evidence
\textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
Thus, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}

@@

\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species. 
Such short forecast horizons are not encouraging.}
\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed}
(Mouquet et al. 2015, Petchey et al. 2015).
\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate}
(Teller et al. 2016),
\textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty}
(Petchey et al. 2015).
\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts}
(Petchey et al. 2015).
\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.

In conclusion, we found that models based on individual-level
demographic data generally failed to generate more skillful population
forecasts than models based on population-level data. This finding runs
counter to our expectations, but is consistent with recent theoretical
(Freckleton et al. 2011) and empirical work (Queenborough et al. 2011).
Moreover, for the two species where including climate covariates
improved forecast skill, both models generated equally skillful
one-step-ahead forecasts. Thus, we conclude that models based on
population-level data, rather than individual-level data, may be
adequate for forecasting the states and dynamics of plant populations.
Unfortunately, our analysis, where climate effects were relatively
unimportant in vital rate regressions, did not allow us to sufficiently
test our prediction that individual-level data is necessary to generate
skillful forecasts if different vital rates respond to climate in
unique, potentially opposing, ways. Nonetheless, our results are
encouraging for the use of easy-to-collect population-level data for
forecasting the state of plant populations.

>>>>

To date, most empirical population modelers have relied on
individual-level data, with few attempts to capitalize on
population-level measures. An important exception was an effort by
Taylor and Hastings (2004) to model the population growth rate of an
invasive species to identify the best strategies for invasion control.
They used a ``density-structured'' model where the state variable is a
discrete density state rather than a continuous density measure. Such
models do not require individual-level demographic data and can
adequately describe population dynamics. Building on Taylor and Hastings
(2004), Freckleton et al. (2011) showed that density-structured models
compare well to continuous models in theory, and Queenborough et al.
(2011) provide empirical evidence that density-structured models are
capable of reproducing population dynamics at landscape spatial scales
(also see Mieszkowska et al. 2013), even if some precision is lost when
compared to fully continuous models. However, previous tests of
density-structured population models have yet to assess their ability to
forecast out-of-sample observations, and they have not included
environmental covariates, which are necessary to forecast population
responses to climate change.

Addressing climate change questions with models fit to population-level
data is potentially problematic. Population growth (or decline) is the
outcome of demographic processes such as survival, growth, and
recruitment that occur at the level of individual plants. Climate can
affect each demographic process in unique, potentially opposing, ways
(Dalgleish et al. 2011). These unique climate responses may be difficult
to resolve in statistical models based on population-level data where
demographic processes are not identifiable. Futhermore, models based on
aggregated data may reflect short-term effects of one vital rate more
than others whose importance may only emerge over the long-term. For
example, a one-year change in a plant species' cover or biomass might
reflect growth or shrinkage of the largest individuals, whereas the
long-term trajectory of the population might be more influenced by
recruitment. The same is true for density dependence: intraspecific
density depedence may act most strongly on vital rates, like
recruitment, that are difficult to identify from population-level data.
If density dependence and/or important climate effects are missed
because of the aggregation inherent in population-level data, then
population models built with such data will make uninformative or
unreliable forecasts.

@@

We fit all models using a hierarchical Bayesian approach. In the
following description, we focus on the main process and the model
likelihood (full model descriptions are in the Supporting Information).
For the likelihood models, \(\textbf{y}^X\) is always the relevant
vector of observations for vital rate \emph{X} (\emph{X} = \emph{S},
\emph{G}, \emph{R}, or \emph{P} for survival, growth, recruitment, or
population growth). For example, \(\textbf{y}^S\) is a vector of 0s and
1s indicating whether a genet survives from \emph{t} to \emph{t+1}, or
not, for all observation years and quadrats. All model parameters are
species-specific, but we omit subscripts for species in model
descriptions below to reduce visual clutter. For brevity, we only
describe models with climate covariates included, but models without
climate covariates are simply the models described below with the
climate effects removed.

@@

Our Bayesian approach to fitting the vital rate models required choosing
appropriate priors for unknown parameters and deciding which, if any, of
those priors should be hierarchical. For each species, we fit yearly
size effects and yearly intercepts hierarchically, where year-specific
coefficients were modeled with global distributions representing the
mean size effect and intercept. Quadrat random effects were also fit
hierarchically, with quadrat offsets modeled using distributions with
mean zero and a shared variance term (independent Gaussian priors).
Climate effects were modeled as independent covariates whose prior
distributions were optimized for prediction using statistical
regularization (see \textbf{Statistical regularization: Bayesian ridge
regression} below).

@@

An important feature of any forecasting model is the rate at which
forecast skill declines as the time between an observation and a
forecast increases.
\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon}
(Petchey et al. 2015). To assess the forecast horizons of our models, we
initiate the forecast model with the population state at some time
\emph{t} and make sequential forecasts of the population at times
\(t+1, t+2, \dots, t+T\) where \emph{T} is the maximum number of years
between the initial year and the final year of our observations. For
example, if we initialize the forecast model with percent cover in 1940,
we are able to make five forecasts up to the year 1945. Forecast models
are not re-initialized with observations between years. Thus, in our
current example, the model forecast for percent cover in 1941 has a
forecast horizon of one year, the forecast in 1942 has a forecast
horizon of two years, and so on. We performed these simulations using
mean parameter values for all model types (IPM with/without climate; QBM
with/without climate) and all possible initial years. For a given
forecast distance, we averaged the correlation between forecasts and
observations. Note that our forecasts for the horizon analysis are all
made using in-sample data because we used model fits from the full data
set. Nonetheless, our simulations offer insight into the differences
among model forecast horizons.
\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold. the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes}
(Petchey et al. 2015),
\textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}

@@

With climate covariates included
\textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased},
but they did so at similar rates (Fig. 4). The only exception is for
\emph{Poa secunda}, where QBM forecast accuracy remained steady as the
\textcolor{blue}{temporal distance of the forecast increased}, whereas
IPM forecast accuracy declined (Fig. 4).
\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H. comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}

@@

Our comparison between a traditional, demographic population model
without environmental forcing (the IPM) and an equivalent model inspired
by density-structured models (the QBM) showed that IPM forecasts of
out-of-sample plant population states were no more accurate than
forecasts from the QBM (Fig. 2; `no-climate' bars). This result differed
from our expectation that the IPM would out-perform the QBM, because of
its mechanistic representation of the perennial life cycle. Our result
also confirms theoretical (Freckleton et al. 2011) and empirical work
(Taylor and Hastings 2004, Queenborough et al. 2011) showing that
density-structured models can be useful surrogates for demographic
models when the goal is to estimate or forecast population states over
large spatial extents.

We also expected the inclusion of environmental forcing to reveal
further differences between the models. Interannual variation in weather
can affect vital rates in different ways (Dalgleish et al. 2011). Thus,
estimates of climate effects on plant population growth may be biased or
non-identifiable when the underlying statistical model is fit using
population-level data that integrates over the potentially unique
climate responses of individual vital rates. We found some evidence that
the QBM failed to detect climate effects for three species (\emph{B.
gracilis}, \emph{H. comata}, and \emph{Poa secunda}), where important
climate effects were identified in the individual vital rate models but
not in the percent cover model (Fig. 3).
\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
Furthermore, for the two species where including climate covariates
increased forecast accuracy (\emph{B. gracilis} and \emph{Poa secunda}),
forecast accuracy (Fig. 2) and error (Fig. S2) were equivalent between
the IPM and QBM.

@@

We also expected IPM forecast accuracy to decline at a lower rate than
the QBM as the
\textcolor{blue}{time between the model initialization and the forecast increased.}
In principle, more mechanistic models should produce better predictions,
especially under novel conditions (Evans 2012, Schindler and Hilborn
2015). In our case, the IPM explicitly models the influence of weather
on recruitment and survival, effects that may be poorly represented in
the QBM because recruitment and survival mainly affect small plants that
contribute little to year-to-year changes in percent cover. Over longer
time scales, the addition and subtraction of small plants could have
large effects on population growth, so explicitly modeling these effects
could contribute to a longer forecast horizon. However, we found no
evidence
\textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}

@@

\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species. 
Such short forecast horizons are not encouraging.}
\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed}
(Mouquet et al. 2015, Petchey et al. 2015).
\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate}
(Teller et al. 2016),
\textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty}
(Petchey et al. 2015).
\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts}
(Petchey et al. 2015).
\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.

In conclusion, we found that models based on individual-level
demographic data generally failed to generate more skillful population
forecasts than models based on population-level data, even in models
which included climate covariates. This finding runs counter to our
expectations, but is consistent with recent theoretical (Freckleton et
al. 2011) and empirical work (Queenborough et al. 2011). We conclude
that models based on population-level data, rather than individual-level
data, may be adequate for forecasting the states and dynamics of plant
populations. This conclusion comes with the caveat that our analysis may
be a weak test of the prediction that individual-level data is necessary
for forecasting if different vital rates respond to climate in opposing
ways, because climate effects were relatively unimportant in our vital
rate regressions. Nonetheless, our results should encourage the use of
easy-to-collect population-level data for forecasting the state of plant
populations.
